---
title: "Biostat 203B Homework 1"
author: Yuyuan Lin
subtitle: Due Jan 21 @ 11:59PM
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Display machine information for reproducibility:
```{r, eval = T}
sessionInfo()
```

## Q1. Git/GitHub

Please refer to my GitHub private repo [biostat-203b-2022-winter](https://github.com/yuyuan-lin/biostat-203b-2022-winter) for the use of `git`.

## Q2. Data ethics training

**My verification links**
    
[Completion Report](https://www.citiprogram.org/verify/?k78527830-9ab5-4fc8-a11f-628b252e0f4b-46576479)
    
[Completion Certificate](https://www.citiprogram.org/verify/?w927396e3-b866-4ccd-a131-82b37ead8898-46576479)
    
If the above links do not work, please copy and paste the following links into your browser:

1. https://www.citiprogram.org/verify/?k78527830-9ab5-4fc8-a11f-628b252e0f4b-46576479
2. https://www.citiprogram.org/verify/?w927396e3-b866-4ccd-a131-82b37ead8898-46576479

## Q3. Linux Shell Commands

1. The `/mnt/mimiciv/1.0` folder on teaching server contains data sets from MIMIC-IV. Refer to the documentation <https://mimic.mit.edu/docs/iv/> for details of data files.  
    ```{bash}
    ls -l /mnt/mimiciv/1.0
    ```

    Use Bash commands to answer following questions.

2. Display the contents in the folders `core`, `hosp`, `icu`. Why are these data files distributed as `.csv.gz` files instead of `.csv` (comma separated values) files? Read the page <https://mimic.mit.edu/docs/iv/> to understand what's in each folder.

    ```{bash}
    ls -l /mnt/mimiciv/1.0/core
    ```

    ```{bash}
    ls -l /mnt/mimiciv/1.0/hosp
    ```

    ```{bash}
    ls -l /mnt/mimiciv/1.0/icu
    ```

    **Answer:** The reason of using `.gz` files: these data files have relatively large size (see above); using `.gz` helps reduce file sizes for transportation or storage purposes. `.gz` files also have handy features like storing original file name and time-stamp, enabling users to recover original file information even after the file was transferred. 
    

3. Briefly describe what bash commands `zcat`, `zless`, `zmore`, and `zgrep` do.

    **Answer:**
    
    - `zcat`: is used to view the content of compressed file; it can also be used to decompress file (combining with pipe).
    - `zless`: is used to paginate a compressed file without reading the entire file, which is same as `zcat [FILENAME] | less`.
    - `zmore`: is also used to view a compressed file page by page, which is same as `zcat [FILENAME] | more`.
    - `zgrep`: is used to search inside a compressed file; it prints lines that match an expression.
    

4. What's the output of following bash script?
    ```{bash, eval=F}
    for datafile in /mnt/mimiciv/1.0/core/*.gz
      do
        ls -l $datafile
      done
    ```

    **Answer:** This bash script shows the information of all the `.gz` data files under the `/mnt/mimiciv/1.0/core/` directory.

    Display the number of lines in each data file using a similar loop.
    ```{bash}
    for datafile in /mnt/mimiciv/1.0/core/*.gz
      do
        zcat $datafile | wc -l
      done
    ```

5. Display the first few lines of `admissions.csv.gz`. How many rows are in this data file? How many unique patients (identified by `subject_id`) are in this data file? (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `sort`, `uniq`, `wc`, and so on.)

    **Answer:**

    Display the first 10 rows:
    ```{bash}
    zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | awk '(NR <= 10)'
    ```

    Display the number of rows in the file:
    ```{bash}
    zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | wc -l
    ```

    Display the number of unique patients:
    ```{bash}
    zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | awk -F, '{if (NR > 1) print $1 }' | sort | uniq | wc -l
    ```

6. What are the possible values taken by each of the variable `admission_type`, `admission_location`, `insurance`, and `ethnicity`? Also report the count for each unique value of these variables. (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `uniq -c`, `wc`, and so on.)

    **Answer:**
    For `admission_type`: there are 9 different possible values, and the counts for each unique values are shown below.
    ```{bash}
    zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | awk -F, '{if (NR > 1) print $6 }' | sort | uniq | wc -l
    ```

    ```{bash}
    zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | awk -F, '{if (NR > 1) print $6 }' | sort | uniq -c
    ```

    For `admission_location`: there are 12 different possible values (including blank/missing), and the counts for each unique values are shown below.
    ```{bash}
    zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | awk -F, '{if (NR > 1) print $7 }' | sort | uniq | wc -l
    ```
    
    ```{bash}
    zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | awk -F, '{if (NR > 1) print $7 }' | sort | uniq -c
    ```

    For `insurance`: there are 3 different possible values, and the counts for each unique values are shown below.
    ```{bash}
    zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | awk -F, '{if (NR > 1) print $9 }' | sort | uniq | wc -l
    ```
    
    ```{bash}
    zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | awk -F, '{if (NR > 1) print $9 }' | sort | uniq -c
    ```

    For `ethnicity`: there are 8 different possible values, and the counts for each unique values are shown below.
    ```{bash}
    zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | awk -F, '{if (NR > 1) print $12 }' | sort | uniq | wc -l
    ```
    
    ```{bash}
    zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | awk -F, '{if (NR > 1) print $12 }' | sort | uniq -c
    ```

## Q4. Who's popular in Price and Prejudice

1. You and your friend just have finished reading *Pride and Prejudice* by Jane Austen. Among the four main characters in the book, Elizabeth, Jane, Lydia, and Darcy, your friend thinks that Darcy was the most mentioned. You, however, are certain it was Elizabeth. Obtain the full text of the novel from <http://www.gutenberg.org/cache/epub/42671/pg42671.txt> and save to your local folder. 
    ```{bash, eval=FALSE}
    wget -nc http://www.gutenberg.org/cache/epub/42671/pg42671.txt
    ```
Explain what `wget -nc` does. Do **not** put this text file `pg42671.txt` in Git. Complete the following loop to tabulate the number of times each of the four characters is mentioned using Linux commands.
    
    **Answer:** `wget -nc` retrieves files from the specified http URL. The flag `-nc` means if the file already existed, `wget` would not download the latest version of the file.

    ```{bash, eval=FALSE}
    wget -nc http://www.gutenberg.org/cache/epub/42671/pg42671.txt
    for char in Elizabeth Jane Lydia Darcy
    do
      echo $char:
      # some commands here
    done
    ```

2. What's the difference between the following two commands?
    ```{bash eval=FALSE}
    echo 'hello, world' > test1.txt
    ```
    and
    ```{bash eval=FALSE}
    echo 'hello, world' >> test2.txt
    ```

3. Using your favorite text editor (e.g., `vi`), type the following and save the file as `middle.sh`:
    ```{bash eval=FALSE}
    #!/bin/sh
    # Select lines from the middle of a file.
    # Usage: bash middle.sh filename end_line num_lines
    head -n "$2" "$1" | tail -n "$3"
    ```
Using `chmod` make the file executable by the owner, and run 
    ```{bash eval=FALSE}
    ./middle.sh pg42671.txt 20 5
    ```
Explain the output. Explain the meaning of `"$1"`, `"$2"`, and `"$3"` in this shell script. Why do we need the first line of the shell script?

### Q5. More fun with Linux

Try following commands in Bash and interpret the results: `cal`, `cal 2021`, `cal 9 1752` (anything unusual?), `date`, `hostname`, `arch`, `uname -a`, `uptime`, `who am i`, `who`, `w`, `id`, `last | head`, `echo {con,pre}{sent,fer}{s,ed}`, `time sleep 5`, `history | tail`.
