---
title: "Biostat 203B Homework 3"
subtitle: Due Feb 25 @ 11:59PM
output:
  # ioslides_presentation: default
  html_document:
    toc: true
    toc_depth: 4
---

Display machine information:
```{r}
sessionInfo()
```

Load database libraries and the tidyverse frontend:
```{r}
suppressPackageStartupMessages(library(dbplyr))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lubridate))
library(DBI)
library(bigrquery)
```

## Q1. Compile the ICU cohort in HW2 from the Google BigQuery database 

Below is an outline of steps.

1. Load the service account token.
```{r}
# path to the service account token 
if (Sys.info()[["nodename"]] == "biostat-203b-teaching-server") {
  # on teaching server
  satoken <- "/mnt/mimiciv/1.0/biostat-203b-2022winter-3fdc2392ac39.json"
} else {
  # on my own Mac laptop
  satoken <- "/Users/yuyuanlin/rcoding/biostat-203b-2022-winter/hw3/biostat-203b-2022winter-3fdc2392ac39.json"
}
# BigQuery authentication using service account
bq_auth(
  path = satoken
)
```

1. Connect to BigQuery database `mimic4_v1_0_203b` in GCP (Google Cloud Platform), using the billing account `biostat-203b-2022winter`.
```{r}
# Connect to the BigQuery database `biostat-203b-2022winter.mimic4_v1_0_203b`
con <- dbConnect(
    bigrquery::bigquery(),
    project = "biostat-203b-2022winter",
    dataset = "mimic4_v1_0_203b",
    billing = "biostat-203b-2022winter"
)
con
```

2. List all tables in the `mimic4_v1_0_203b` database.
```{r}
dbListTables(con)
```

3. Connect to the `icustays` table.
```{r}
# full ICU stays table
icustays_tble <- tbl(con, "icustays") %>%
  show_query() %>%
  print(width = Inf)
```
We only keep the first ICU stay. Following code is kind of a hack, using the `summarise_all(min)` function. It seems that `slice_min()`, `slice_head()`, `distinct(, .keep_all = TRUE)` don't work with `dbplyr`+`bigrquery` at the moment.
```{r}
icustays_tble <- icustays_tble %>% 
  select(subject_id, intime) %>% 
  group_by(subject_id) %>% 
  summarise_all(min) %>%
  left_join(icustays_tble, by = c("subject_id", "intime")) %>%
  show_query() %>%
  print(width = Inf)
```


4. Connect to the `admissions` table and only keep the patients who have a match in `icustays_tble` (according to `subject_id` and `hadm_id`).
```{r}
admissions_tble <- tbl(con, "admissions") %>%
  semi_join(icustays_tble, by = c("subject_id", "hadm_id")) %>%
  print(width = Inf)
```

5. Connect to the `patients` table and only keep the patients who have a match in `icustays_tble` (according to `subject_id`).
```{r}
patients_tble <- tbl(con, "patients") %>%
  semi_join(icustays_tble, by = c("subject_id")) %>%
  print(width = Inf)
```

6. Connect to the `labevents` table and retrieve a subset that only contain subjects who appear in `icustays_tble` and the lab items listed in HW2. 
```{r}
labevents_tble <- tbl(con, "labevents") %>%
  select(c("subject_id", "itemid", "charttime", "valuenum")) %>%
  semi_join(icustays_tble, by = c("subject_id")) %>%
  filter(itemid %in% c(50912, 50971, 50983, 50902, 50882, 
                       51221, 51301, 50931, 50960, 50893)) %>%
  print(width = Inf)
```
Only keep the first lab measurements during ICU stay and pivot lab items to become variables/columns.
```{r}
labevents_tble_w <- labevents_tble %>% 
  # remove lab events outside the specific ICU stay
  left_join(icustays_tble) %>%
  filter(charttime > intime & charttime < outtime) %>%
  select(colnames(labevents_tble)) %>%
  # pick the first item for each patients
  select(subject_id, charttime, itemid) %>% 
  group_by(subject_id, itemid) %>% 
  summarise_all(min) %>%
  left_join(labevents_tble, by = c("subject_id", "charttime", "itemid")) %>%
  # pivot to wider
  select(-c("charttime")) %>%
  pivot_wider(names_from = itemid, names_prefix = "itemid_",
              values_from = valuenum) %>%
  # show results
  print(width = Inf)
```

7. Connect to `chartevents` table and retrieve a subset that only contain subjects who appear in `icustays_tble` and the chart events listed in HW2.
```{r}
chartevents_tble <- tbl(con, "chartevents") %>%
  select(c("subject_id", "itemid", "charttime", "valuenum")) %>%
  semi_join(icustays_tble, by = c("subject_id")) %>%
  filter(itemid %in% c(220045, 220181, 220179, 223761, 220210)) %>%
  print(width = Inf)
```
Only keep the first chart events during ICU stay and pivot chart events to become variables/columns.
```{r}
chartevents_tble_w <- chartevents_tble %>% 
  # remove lab events outside the specific ICU stay
  left_join(icustays_tble) %>%
  filter(charttime > intime & charttime < outtime) %>%
  select(colnames(chartevents_tble)) %>%
  # pick the first item for each patients
  select(subject_id, charttime, itemid) %>% 
  group_by(subject_id, itemid) %>% 
  summarise_all(min) %>%
  left_join(chartevents_tble, by = c("subject_id", "charttime", "itemid")) %>%
  # pivot to wider
  select(-c("charttime")) %>%
  pivot_wider(names_from = itemid, names_prefix = "itemid_",
              values_from = valuenum) %>%
  # show results
  print(width = Inf)
```

8. Put things together. This step is similar to Q7 of HW2. Using one chain of pipes `%>%` to perform following data wrangling steps: (i) start with the `icustays_tble` for the first ICU stay of each unique patient, (ii) merge in admissions and patients tables, (iii) keep adults only (age at admission >= 18), (iv) merge in the labevents and chartevents tables, (v) create an indicator for 30-day mortality, (vi) save the final tibble to an `icu_cohort.rds` R data file in the `mimiciv_shiny` folder.

```{r}
# make a directory mimiciv_shiny
if (!dir.exists("mimiciv_shiny")) {
  dir.create("mimiciv_shiny")
}
```

```{r}
icu_cohort <- icustays_tble %>% 
  # join admissions table
  left_join(admissions_tble, by = c("subject_id", "hadm_id")) %>%
  # join patients table
  left_join(patients_tble, by = c("subject_id")) %>%
  # keep adults only
  mutate(admit_age = year(admittime) - anchor_year + anchor_age) %>%
  filter(admit_age >= 18) %>%
  # join lab event table (wide)
  left_join(labevents_tble_w, by = c("subject_id")) %>%
  # join chart event table (wide)
  left_join(chartevents_tble_w, by = c("subject_id")) %>%
  # create 30-day-mortality indicator
  mutate(
    thirty_day_mort = case_when(is.na(dod) ~ 0,
                                as.Date(admittime) + 30L > dod ~ 1,
                                as.Date(admittime) + 30L < dod ~ 0)
  ) %>%
  # print results
  print(width = Inf)
```

```{r}
# write tbl_dbi to local rds file
icu_cohort %>% 
  collect() %>%
  saveRDS(file = "mimiciv_shiny/icu_cohort.rds")
```

Close database connection and clear workspace.
```{r, eval=F}
dbDisconnect(con)
rm(list = ls())
```

## Q2. Shiny app

Develop a Shiny app for exploring the ICU cohort data created in Q1. The app should reside in the `mimiciv_shiny` folder. The app should provide easy access to the graphical and numerical summaries of variables (demographics, lab measurements, vitals) in the ICU cohort.


<p style="color:red;">For this part, please refer to the `app.r` located in the `mimiciv_shiny` folder. </p>